---
title: "Getting Started with ibaq4diann"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ibaq4diann}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Introduction

**ibaq4diann** computes intensity-Based Absolute Quantification (iBAQ) values
from DIA-NN search results and a reference proteome FASTA. iBAQ normalises
summed precursor intensities by the number of theoretically observable tryptic
peptides, making protein abundances comparable across proteins of different
sizes.

$$iBAQ = \frac{\sum \text{precursor intensities}}{\text{# theoretical tryptic peptides}}$$

This vignette demonstrates the full workflow using a minimal toy example
(simulated data) to illustrate the API without requiring large input files.

## Installation

```{r install, eval = FALSE}
# Bioconductor dependency (if not already installed)
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("Biostrings")

# Install ibaq4diann from GitHub
remotes::install_github("AlisonChaves/ibaq4diann")
```

```{r setup}
library(ibaq4diann)
```

## Step 1: Prepare inputs

`compute_ibaq()` requires two inputs:

1. A **DIA-NN `report.parquet`** file — the main output of a DIA-NN
   database search.
2. A **reference proteome FASTA** — the same database used during the search
   (e.g. a UniProt proteome download).

```{r inputs, eval = FALSE}
parquet_path <- "report.parquet"
fasta_path   <- "UP000005640_9606.fasta"  # Human UniProt proteome
```

## Step 2: Compute iBAQ values

```{r compute, eval = FALSE}
ibaq <- compute_ibaq(
  parquet_path = parquet_path,
  fasta_path   = fasta_path
)

# Wide format: proteins × samples
head(ibaq)
dim(ibaq)
```

The returned tibble has:

- `protein_id` — protein group identifier (from DIA-NN `Protein.Group`)
- `Protein.Names` — full protein name (if present in the report)
- `Genes` — gene name(s) (if present)
- `n_theoretical_peptides` — in-silico tryptic peptide count (iBAQ denominator)
- One column per DIA-NN run containing the iBAQ value (log2-transformed by
  default)

## Step 3: Customise parameters

```{r custom, eval = FALSE}
# Raw (non-log2) iBAQ, long-format output, allow 1 missed cleavage
ibaq_custom <- compute_ibaq(
  parquet_path         = parquet_path,
  fasta_path           = fasta_path,
  intensity_col        = "Precursor.Quantity",  # raw intensities
  q_value_cutoff       = 0.01,
  lfq_quality_cutoff   = NULL,                  # skip LFQ quality filter
  max_missed_cleavages = 1,
  log2_transform       = FALSE,
  output_long          = TRUE                   # tidy long format
)

head(ibaq_custom)
```

## Step 4: Save results

```{r save, eval = FALSE}
# Write to TSV
readr::write_tsv(ibaq, "iBAQ_results.tsv")

# Or use the convenience pipeline wrapper
run_ibaq_pipeline(
  parquet_path = parquet_path,
  fasta_path   = fasta_path,
  output_path  = "iBAQ_results.tsv"
)
```

## Step 5: Visualise results

`ibaq4diann` provides four ready-made plotting functions, all returning
`ggplot2` objects that can be further customised.

### Ranked protein abundances

```{r rank_plot, eval = FALSE}
p_rank <- plot_ibaq_rank(
  ibaq,
  top_n_labels = 10,       # label top 10 proteins per sample
  label_col    = "Genes"   # use gene names as labels
)
print(p_rank)
ggplot2::ggsave("ibaq_rank.pdf", p_rank, width = 12, height = 5)
```

Each facet shows one sample. The x-axis is the protein rank (rank 1 = most
abundant); the y-axis shows log₂(iBAQ + 1).

### iBAQ value distributions

```{r dist_plot, eval = FALSE}
p_dist <- plot_ibaq_distribution(
  ibaq,
  plot_type = "both"   # violin + box plot overlay
)
print(p_dist)
```

Useful for QC: samples with similar distribution shapes and medians indicate
good between-sample reproducibility.

### Sample correlation heatmap

```{r cor_plot, eval = FALSE}
p_cor <- plot_ibaq_correlation(
  ibaq,
  method = "pearson"
)
print(p_cor)
```

High pairwise correlations (> 0.95) between replicates confirm experimental
reproducibility.

### Dynamic range

```{r dr_plot, eval = FALSE}
p_dr <- plot_ibaq_dynamic_range(ibaq)
print(p_dr)
```

Each vertical segment represents the full iBAQ value range of a sample; the
diamond marks the median. The number of quantified proteins (`n=`) is annotated
at the top of each bar.

## Key parameters reference

| Parameter | Default | Description |
|---|---|---|
| `intensity_col` | `"Precursor.Normalised"` | Intensity column to use |
| `q_value_cutoff` | `0.01` | Precursor FDR threshold |
| `pg_q_value_cutoff` | `0.01` | Protein group FDR threshold |
| `proteotypic_only` | `TRUE` | Keep only unique peptides |
| `lfq_quality_cutoff` | `0.50` | Minimum LFQ quality score |
| `min_peptide_len` | `6` | Minimum tryptic peptide length |
| `max_peptide_len` | `30` | Maximum tryptic peptide length |
| `max_missed_cleavages` | `0` | Missed cleavages in silico |
| `log2_transform` | `TRUE` | Apply log₂(x + 1) transform |
| `output_long` | `FALSE` | Return long vs. wide format |

## References

- Schwanhäusser B et al. (2011). Global quantification of mammalian gene
  expression control. *Nature* **473**:337–342.
  [doi:10.1038/nature10098](https://doi.org/10.1038/nature10098)
- DIA-NN: [github.com/vdemichev/DiaNN](https://github.com/vdemichev/DiaNN)
